---
title: "Lead Exposure and IQ Loss in Children: a Bayesian and Frequentist Meta Analysis of Recent Epidemiological Studies"
author:
  - name: Paulina Sell
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0009-0007-7419-2956
  - name: Margaux Sanchez
    affiliation: Agence nationale de sécurité sanitaire
    orcid: 
  - name: Philippe Palmont
    affiliation: Agence nationale de sécurité sanitaire
    orcid: 
  - name: Dietrich Plass
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0000-0002-8058-1399
date-modified: "2026-01-12"
bibliography: references.bib
csl: vancouver.csl
reference-section-title: "References"
format:
  html:
    toc: true
    title-block-banner: false
    number-sections: true
    self-contained: true
    embed-resources: true
    theme: cosmo
    page-layout: full
comments:
  hypothesis: true
  # docx:
  #   reference-doc: ../manuscript/ref.docx
  #   number-sections: true
  #   toc: true
  #   toc-depth: 5
    # fig-align: "center"
    # fig-width: 6.5
    # fig-height: 3.5
    # fig-dpi: 600
    # out-width: "40%"
---

```{css}
#| echo: false
p {
  text-align: justify
}
```

# Notes {.unnumbered}

-   Journal: [Population Health Metrics, BMC](https://link.springer.com/journal/12963/submission-guidelines#aimsAndScope)
-   Using [PRISMA 2020 checklist](https://www.prisma-statement.org/prisma-2020) for structure
-   Using [raRoB tool](https://efsa.onlinelibrary.wiley.com/doi/10.2903/sp.efsa.2024.EN-8925) for Risk of Bias Assessment

# Abstract {.unnumbered}

# Background

-   [x] Intro: Lead, health effects, burden of disease
-   [x] Currently used ERFs
-   [ ] Rationale for methodological comparison (BMA - FMA)
-   [x] Aim of this study\

Lead (Pb) is a pervasive environmental pollutant originating from anthropogenic and natural sources. 
Although concentrations have declined in past decades following the phase-out of leaded gasoline, lead remains present in the environment and human blood [@salamanca-fernandez2025SociodemographicDeterminantsTemporal; @lacerda2023GlobalDecreaseBlood].
The main exposure pathways in Europe are dietary products, drinking water, and indoor dust [@europeanfoodsafetyauthority2012LeadDietaryExposure; @sy2024AssessmentLongTermExposure].
The Global Burden of Disease Study 2023 estimated that 3.4 Million	(95% Confidence Interval (CI): 2.7; 4.0 Million) Disability-Adjusted Life Years (DALYs) were attributable to lead exposure in 2023 in the European Union (EU) @brauer2024GlobalBurdenStrength.

Among the various adverse health outcomes associated with low lead exposure — including
infertility in women @chang2006LowBloodLead, 
attention-deficit/hyperactivity disorder in children @dimitrov2024SystematicReviewMetaanalysis , 
antisocial behavior @shaffer2025LeadExposureAntisocial, 
decreased kidney function @harari2018BloodLeadLevels, 
liver disease @wan2022ChronicLeadExposure, 
coronary artery atherosclerosis @rosengren2025ExposureLeadCoronary and 
cardiovascular mortality @lanphear2018LowlevelLeadExposure,
impaired intellectual development in children represents a particularly critical endpoint for public health policy @crump2013StatisticalReevaluationData.
The effects of lead on the IQ score (intelligence quotient) are lifelong and therefore lead to substantial societal costs @larsen2023GlobalHealthBurden.
Children are especially vulnerable to lead neurotoxicity due to their developing nervous systems and relatively high exposure per body mass.
Importantly, there is no evidence for a threshold concentration below which lead's detrimental effects on the intellectual development can be assumed to be negligible @efsapaneloncontaminantsinthefoodchaincontam2010ScientificOpinionLead.

It is essential to continue to accurately characterize the exposure-response relationship of lead exposure and IQ loss not only because lead concentrations today are lower than at the time when previous studies were conducted, but also because of substantive criticism concerning the existing evidence base. \
Current burden of disease assessments on lead-related IQ loss at national and global scales [@brauer2024GlobalBurdenStrength; @carrington2019GlobalBurdenIntellectual; @figueroa2024LossCognitiveFunction; @larsen2023GlobalHealthBurden; @mcfarland2022HalfUSPopulation] predominantly rely on exposure-response functions (ERFs) from two studies: the pooled analysis by Lanphear et al. (2005) or its subsequent re-analysis by Crump et al. (2013) [@crump2013StatisticalReevaluationData; @lanphear2005LowLevelEnvironmentalLead].
While these are important landmark studies shaping the course of research on health effects of lead exposure in children, methodological issues and errors in the Lanphear et al. 2005 article have been raised [@lanphear2019ErratumLowLevelEnvironmental; @crump2013StatisticalReevaluationData] and the stepwise backward-elimination covariate selection approach employed by Crump et al. (2013), which is central to their findings, has received substantial criticism [@thompson1995StepwiseRegressionStepwise; @harrell2015RegressionModelingStrategies; @smith2018StepAwayStepwise].
These limitations warrant reconsideration of their continued use as evidence base.

Another motivation to conduct an updated meta analysis today, is that while Lanphear et al. investigated low BLL (<10 µg/dL) exposures in their study, BLL in the EU have decreased even further and now mostly lie below 5 µg/dL.
In order to provide more relevant risk estimates, a synthesis of new epidemiological studies is needed.
The pooled exposure-response relationship can then be applied to risk assessments and used by policymakers. \
Beyond updating the evidence base, there is also a methodological opportunity to enhance the utility of meta-analytic findings.
While frequentist meta-analysis has been a conventional approach in environmental epidemiology, Bayesian meta-analysis offers distinct advantages.
Specifically, the posterior distributions generated by Bayesian approaches offer consistent and holistic (?) quantification of uncertainty that can be directly incorporated into probabilistic Burden of Disease studies.
As probabilistic methods become increasingly adopted in contemporary risk assessment and policy evaluation, having exposure-response estimates in a format that integrates seamlessly with these frameworks represents a practical advancement beyond purely methodological considerations.

The aim of this study is to summarize and quantitatively pool the available epidemiological literature on childrens blood lead levels and neurodevelopment (operationalized as IQ score).
We conduct both a frequentist meta-analysis—representing the established approach widely used in epidemiology—and compare it to a Bayesian meta-analysis, evaluating how methodological choices influence risk estimates and their applicability to probabilistic burden of disease assessment.

# Methods

-   [ ] Overview: systematic review and meta analysis, comparing a frequentist to a Bayesian approach
-   [ ] Inclusion & exclusion criteria, databases, search strategy, screening process
-   [ ] Risk of bias analysis method
-   [ ] Description & rationale of harmonization of reported effect measures
-   [ ] Description of meta analysis methodology (frequentist & Bayesian), models and software used, heterogeneity identification
-   [ ] Sensitivity analysis\

In order to quantitatively synthesize the available epidemiological literature on the association of childrens blood lead levels and their IQ, we conducted a systematic review and meta analysis. 
We compare a frequentist meta analysis approach to a Bayesian approach that has specific advantages and has gained popularity in recent years but is still not widely used in the field of epidemiology.\
(Information on literature review)\
We conducted a Risk of Bias assessment using a rapid Risk of Bias (raROB) tool with a specific focus on human observational epidemiological studies @plate2024DevelopmentToolRapid. 
Each included publications risk of bias was appraised by answering 15 items in the domains selection, exposure, outcome, confounding, follow-up, analysis, and selective reporting. 
The overall risk of bias for each individual study is then categorized as either low, moderate or high. 
Each publication was first assessed by one researcher and then consensus with a large language model (LLM) was sought. 
This was done by uploading the publications and the list of all raRoB items. 
If there were any deviations in the judgement of the researcher and the LLM, the corresponding item was revisited by the researcher and they made the final decision on each item.

# Results

-   [ ] Cite studies that might appear to meet inclusion criteria, but were excluded & explain why
-   [ ] Study characteristics (table), including original and harmonized effect measures
-   [ ] Flow diagram, description of (fulltext) screening results
-   [ ] RoB results for each included study
-   [ ] Forest plots
-   [ ] Sensitivity analysis results

# Discussion

-   [ ] General interpretation of results in light of evidence in literature
-   [ ] Discuss differences, similarities, strenghts and weaknesses of freq vs. Bayesian approach
-   [ ] Study limitations & strengths

# Conclusions

-   [ ] Conclusion, with reference to aim of study
-   [ ] Implications of results: policy, (scientific?) practice, future research needs

# List of Abbreviations {.unnumbered}

```{r}
#| label: table example
#| echo: false

Meaning <- c("Environmental Burden of Disease", "Years of Life Lost", "Years Lived with Disability", "Disability-adjusted Life Year")
Abbreviation <- c("EBD", "YLL", "YLD", "DALY")
table_abbr <- data.frame(Abbreviation, Meaning)

knitr::kable(table_abbr)
```

# Funding {.unnumbered}

This work was carried out in the framework of the European Partnership for the Assessment of Risks from Chemicals (PARC) and has received funding from the European Union’s Horizon Europe research and innovation programme under Grant Agreement No 101057014. 
Views and opinions expressed are however those of the author(s) only and do not necessarily reflect those of the European Union or the Health and Digital Executive Agency. 
Neither the European Union nor the granting authority can be held responsible for them.

```{r}
#| label: test_ref
#| include: false


#m.brm <- readRDS("../models/main/m.brm_full.rds")
#fitPrior <- readRDS("../models/main/fitPrior_full.rds")
```

```{r}
#| label: test
#| echo: false
#| include: false


#summary(m.brm)
```
