---
title: "Lead Exposure and IQ Loss: Comparison of a Bayesian vs. Frequentist Meta Analysis Approach"
author:
  - name: Paulina Sell
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0009-0007-7419-2956
  - name: Margaux Sanchez
    affiliation: Agence nationale de sécurité sanitaire
    orcid: 
  - name: Philippe Palmont
    affiliation: Agence nationale de sécurité sanitaire
    orcid: 
  - name: Dietrich Plass
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0000-0002-8058-1399
date-modified: "2025-12-15"
bibliography: references.bib
csl: vancouver.csl  # citation style
reference-section-title: "References"
format:
  html:
    toc: true
    title-block-banner: true
    number-sections: true
    self-contained: true
    embed-resources: true
    theme: cosmo
    page-layout: full
  # docx:
  #   reference-doc: ../manuscript/ref.docx
  #   number-sections: true
  #   toc: true
  #   toc-depth: 5
    # fig-align: "center"
    # fig-width: 6.5
    # fig-height: 3.5
    # fig-dpi: 600
    # out-width: "40%"
---

```{r}
#| label: test_ref
#| include: false


m.brm <- readRDS("../models/m.brm_full.rds")
fitPrior <- readRDS("../models/fitPrior_full.rds")
```

```{r}
#| label: test
#| echo: false
#| include: false


summary(m.brm)
```

# Notes {.unnumbered}
* Journal: [Population Health Metrics, BMC](https://link.springer.com/journal/12963/submission-guidelines#aimsAndScope)
* Using [PRISMA 2020 checklist](https://www.prisma-statement.org/prisma-2020) for structure
* Using [raRoB tool](https://efsa.onlinelibrary.wiley.com/doi/10.2903/sp.efsa.2024.EN-8925) for Risk of Bias Assessment
* Test reference @cao2010DoesBackgroundPostnatal 



# Abstract {.unnumbered}

# Background
- [ ] Short intro: Lead, health effects, burden of disease, exposure in Germany
- [ ] Aim of this study

# Methods
- [ ] Overview: systematic review and meta analysis, comparing a frequentist to a Bayesian approach
- [ ] Inclusion & exclusion criteria, databases, search strategy, screening process
- [ ] (Risk of bias analysis method)
- [ ] Description & rationale of harmonization of reported effect measures
- [ ] Description of meta analysis methodology (frequentist & Bayesian), models and software used, heterogeneity identification
- [ ] Method to assess robustness of synthesized results

# Results
- [ ] Cite studies that might appear to meet inclusion criteria, but were excluded & explain why
- [ ] Study characteristics (table), including original and harmonized effect measures
- [ ] Flow diagram, description of (fulltext) screening results
- [ ] RoB results for each included study
- [ ] Forest plots

# Discussion
- [ ] General interpretation of results in light of evidence in literature
- [ ] Discuss differences, similarities, strenghts and weaknesses of freq vs. Bayesian approach
- [ ] Study limitations & strengths

# Conclusions
- [ ] Conclusion, with reference to aim of study
- [ ] Implications of results: policy, (scientific?) practice, future research needs

# List of Abbreviations {.unnumbered}
```{r}
#| label: table example
#| echo: false

Meaning <- c("Environmental Burden of Disease", "Years of Life Lost", "Years Lived with Disability", "Disability-adjusted Life Year")
Abbreviation <- c("EBD", "YLL", "YLD", "DALY")
table_abbr <- data.frame(Abbreviation, Meaning)

knitr::kable(table_abbr)
```

