---
title: "Lead Exposure and IQ Loss in Children: a Bayesian and Frequentist Meta Analysis of Recent Epidemiological Studies"
author:
  - name: Paulina Sell
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0009-0007-7419-2956
  - name: Margaux Sanchez
    affiliation: Agence nationale de sécurité sanitaire
    orcid: 
  - name: Philippe Palmont
    affiliation: Agence nationale de sécurité sanitaire
    orcid: 
  - name: Dietrich Plass
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0000-0002-8058-1399
date-modified: "2026-01-13"
bibliography: references.bib
csl: vancouver.csl
reference-section-title: "References"
format:
  html:
    toc: true
    title-block-banner: false
    number-sections: true
    self-contained: true
    embed-resources: true
    theme: cosmo
    page-layout: full
  # docx:
  #   reference-doc: ../manuscript/ref.docx
  #   number-sections: true
  #   fig-align: "center"
  #   fig-width: 6.5
  #   fig-height: 3.5
  #   fig-dpi: 600
  #   out-width: "40%"
---

```{css}
#| echo: false
p {
  text-align: justify
}
```

# Notes {.unnumbered}

-   Journal: [Population Health Metrics, BMC](https://link.springer.com/journal/12963/submission-guidelines#aimsAndScope)
-   Using [PRISMA 2020 checklist](https://www.prisma-statement.org/prisma-2020) for structure
-   Using [raRoB tool](https://efsa.onlinelibrary.wiley.com/doi/10.2903/sp.efsa.2024.EN-8925) for Risk of Bias Assessment

# Abstract {.unnumbered}

\newpage

# Background

-   [x] Intro: Lead, health effects, burden of disease
-   [x] Currently used ERFs
-   [x] Rationale for methodological comparison (BMA - FMA)
-   [x] Aim of this study\

Lead (Pb) is a pervasive environmental pollutant originating from anthropogenic and natural sources. 
Although concentrations have declined in past decades following the phase-out of leaded gasoline, lead remains present in the environment and human blood [@salamanca-fernandez2025SociodemographicDeterminantsTemporal; @lacerda2023GlobalDecreaseBlood].
In Europe, primary exposure pathways include dietary products, drinking water, and indoor dust [@europeanfoodsafetyauthority2012LeadDietaryExposure; @sy2024AssessmentLongTermExposure].
The Global Burden of Disease Study 2023 estimated that 3.4 million	(95% Confidence Interval (CI): 2.7; 4.0 million) Disability-Adjusted Life Years (DALYs) were attributable to lead exposure in 2023 in the European Union (EU) @brauer2024GlobalBurdenStrength, underscoring the continued public health significance even of low-level exposure.

Lead exposure is associated with multiple adverse health outcomes across the lifespan, including
decreased kidney function @harari2018BloodLeadLevels, 
liver disease @wan2022ChronicLeadExposure, 
coronary artery atherosclerosis @rosengren2025ExposureLeadCoronary,
cardiovascular mortality @lanphear2018LowlevelLeadExposure,
infertility in women @chang2006LowBloodLead, 
attention-deficit/hyperactivity disorder in children @dimitrov2024SystematicReviewMetaanalysis, and 
antisocial behavior @shaffer2025LeadExposureAntisocial.
Among these endpoints, impaired cognitive development in children — manifested as reduced IQ scores — @crump2013StatisticalReevaluationData represents a particularly critical endpoint for public health policy.
Children are especially vulnerable to lead's neurotoxicity due to their developing nervous systems and relatively high exposure per body mass.
Because lead-associated IQ decrements are irreversible and persist throughout life, they lead to substantial and enduring societal costs @larsen2023GlobalHealthBurden, making continued efforts to accurately characterize the exposure-response relationship important for evidence-based policy.

Contemporary blood lead levels (BLL) in the EU have declined markedly and now predominantly fall below 5 µg/dL ([HBM4EU Dashboard](https://hbm.vito.be/eu-hbm-dashboard)), well below the concentrations examined in many earlier studies (<10 µg/dL).
Importantly, there is no evidence for a threshold concentration below which the effects of lead on the IQ can be assumed to be negligible @efsapaneloncontaminantsinthefoodchaincontam2010ScientificOpinionLead
There is evidence for a supralinear exposure-response relationship with steeper IQ decrements per unit increase in BLL at low concentrations compared to higher concentration ranges [@canfield2003IntellectualImpairmentChildren; @lanphear2005LowLevelEnvironmentalLead].
These findings highlight the need for an updated quantitative synthesis that reflects both recent epidemiological evidence and the lower exposure range now predominant in European populations, incorporating contemporary data to inform risk assessment at currently relevant exposure levels.

Existing burden of disease assessments at national and global scales [@brauer2024GlobalBurdenStrength; @carrington2019GlobalBurdenIntellectual; @figueroa2024LossCognitiveFunction; @larsen2023GlobalHealthBurden; @mcfarland2022HalfUSPopulation] rely predominantly on two foundational studies: a pooled analysis of seven cohort-studies by Lanphear et al. (2005) or its subsequent re-analysis by Crump et al. (2013) [@crump2013StatisticalReevaluationData; @lanphear2005LowLevelEnvironmentalLead].
While these landmark studies have substantially shaped research and policy on childhood lead exposure, methodological issues and errors in the Lanphear et al. 2005 article have been raised [@lanphear2019ErratumLowLevelEnvironmental; @crump2013StatisticalReevaluationData] and the stepwise backward-elimination regression approach used by Crump et al. (2013), which is central to their findings, has received substantial criticism [@thompson1995StepwiseRegressionStepwise; @harrell2015RegressionModelingStrategies; @smith2018StepAwayStepwise].
These limitations, combined with the availability of more recent epidemiological data, underscore the need not only for updated evidence synthesis but also for (methodologically) robust  approaches that can provide reliable exposure-response estimates for burden of disease and risk assessments.

Methodological advances in meta-analysis offer an opportunity to enhance the utility of synthesized estimates for burden of disease and risk assessment applications. 
While frequentist meta-analysis has been a conventional approach in environmental epidemiology, Bayesian meta-analysis provides a coherent framework for uncertainty quantification through posterior distributions that inherently incorporate both within-study and between-study variability @grant2025BayesianMetaAnalysisPractical.
Unlike frequentist confidence intervals, which are often misinterpreted in probabilistic terms, Bayesian credible intervals directly represent the probability distribution of the parameter of interest, facilitating transparent communication of uncertainty.
Furthermore, Bayesian posterior distributions can be directly propagated into probabilistic burden of disease studies and risk assessments, providing seamless integration with the probabilistic frameworks increasingly adopted in public health decision-making.
By comparing frequentist and Bayesian approaches, we will explore how inferential framework and uncertainty quantification influence exposure-response estimates and their application.

The aim of this study is twofold: first, to synthesize and quantitatively pool available epidemiological findings on the association between blood lead levels and IQ scores in children; and second, to conduct both a frequentist and Bayesian meta-analyses, comparing how these analytical frameworks influence exposure-response estimates, uncertainty characterization, and applicability to burden of disease and risk assessment.


# Methods

-   [ ] Overview: systematic review and meta analysis, comparing a frequentist to a Bayesian approach
-   [ ] Inclusion & exclusion criteria, databases, search strategy, screening process
-   [ ] Risk of bias analysis method
-   [ ] Description & rationale of harmonization of reported effect measures
-   [ ] Description of meta analysis methodology (frequentist & Bayesian), models and software used, heterogeneity identification
-   [ ] Sensitivity analysis\

In order to quantitatively synthesize the available epidemiological literature on the association of children's blood lead levels and their IQ, we conducted a systematic review and meta analysis. 
We compare a frequentist meta analysis approach to a Bayesian approach that has specific advantages and has gained popularity in recent years but is still not widely used in the field of epidemiology.\
(Section on literature review)\
We conducted a Risk of Bias assessment using a rapid Risk of Bias (raROB) tool with a specific focus on human observational epidemiological studies @plate2024DevelopmentToolRapid. 
Each included publications risk of bias was appraised by answering 15 items in the domains selection, exposure, outcome, confounding, follow-up, analysis, and selective reporting. 
The overall risk of bias for each individual study was then categorized as either low, moderate or high. 
Each publication was first assessed by one researcher and then consensus with a large language model (LLM) was sought. 
This was done by uploading the publications and the list of all raRoB items. 
If there were any deviations in the judgement of the researcher and the LLM, the corresponding item was revisited by the researcher and they made the final decision on the respective item.

# Results

-   [ ] Cite studies that might appear to meet inclusion criteria, but were excluded & explain why
-   [ ] Study characteristics (table), including original and harmonized effect measures
-   [ ] Flow diagram, description of (full text) screening results
-   [ ] RoB results for each included study
-   [ ] Forest plots
-   [ ] Sensitivity analysis results

# Discussion

-   [ ] General interpretation of results in light of evidence in literature
-   [ ] Discuss differences, similarities, strengths and weaknesses of freq vs. Bayesian approach
-   [ ] Study limitations & strengths

# Conclusions

-   [ ] Conclusion, with reference to aim of study
-   [ ] Implications of results: policy, (scientific?) practice, future research needs

# List of Abbreviations {.unnumbered}

```{r}
#| label: table example
#| echo: false
#| classes: plain

Meaning <- c("Blood lead levels", "Confidence interval", "Environmental Burden of Disease", "European Union", "Intelligence quotient", "Years of Life Lost", "Years Lived with Disability", "Disability-adjusted Life Year")
Abbreviation <- c("BLL", "CI", "EBD", "EU", "IQ", "YLL", "YLD", "DALY")
table_abbr <- data.frame(Abbreviation, Meaning)

knitr::kable(table_abbr)
```

# Funding {.unnumbered}

This work was carried out in the framework of the European Partnership for the Assessment of Risks from Chemicals (PARC) and has received funding from the European Union’s Horizon Europe research and innovation programme under Grant Agreement No 101057014. 
Views and opinions expressed are however those of the author(s) only and do not necessarily reflect those of the European Union or the Health and Digital Executive Agency. 
Neither the European Union nor the granting authority can be held responsible for them.

```{r}
#| label: test_ref
#| include: false


#m.brm <- readRDS("../models/main/m.brm_full.rds")
#fitPrior <- readRDS("../models/main/fitPrior_full.rds")
```

```{r}
#| label: test
#| echo: false
#| include: false


#summary(m.brm)
```
