---
title: "Abstract ISEE 2026"
author:
  - name: Paulina Sell
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0009-0007-7419-2956
  - name: Margaux Sanchez
    affiliation: Agence nationale de sécurité sanitaire, Maisons-Alfort, France
    orcid: 
  - name: Philippe Palmont
    affiliation: Agence nationale de sécurité sanitaire, Maisons-Alfort, France
    orcid: 
  - name: Dietrich Plass
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0000-0002-8058-1399
date-modified: last-modified
date-format: "D MMMM YYYY"
bibliography: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/references.bib
csl: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/vancouver.csl
reference-section-title: "References"
format:
    html:
        toc: true
        title-block-banner: false
        number-sections: false
        embed-resources: true
        css: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/custom.scss
        page-layout: full
        theme: yeti
execute:
  cache: false
---

## Information for Authors:
* Deadline: February 27, 2026 (23:59 CET)
* The submitting author is required to ensure that all co-authors are aware of the content of the abstract before submission
* Use acronyms only when necessary and define all non-standard abbreviations and concepts in your abstract at first use
* Abstract titles can contain a maximum of **20 words** (no abbreviations)
* The abstract body should not exceed **300 words** and should consist of the four sections: Objective, Material and Methods, Results, Conclusion.
* Phrases such as “results will be discussed” or “data will be presented” should not be used.
* Please mention all sources of funding for the work described and other acknowledgements briefly at the end of the abstract text.

* A main topic that is relevant for your abstract must be selected during the submission process. 
The Scientific Organizing Committee reserves the right to change the topics during the abstract review process.
* Please enter at least 2 to 5 keywords from the official keyword list

* Topic: A-10: Chemical exposures / A-18: Methods ?
* Keywords: Chemical exposures, Environmental epidemiology, Heavy metals, Lead, International collaboration, Modeling

```{r}
#| label: read_model
#| echo: false

m.brm <- readRDS("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/models/main/m.brm_full.rds")
freq_model <- readRDS("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/models/main/freq_full.rds")
```

```{r}
#| label: extract_results
#| echo: false
#| message: false


library(posterior)
library(HDInterval)

freq_mean <- round(as.numeric(freq_model$beta), 2)
freq_lb <- round(as.numeric(freq_model$ci.lb), 2)
freq_ub <- round(as.numeric(freq_model$ci.ub), 2)

m.brm_samples <- as_draws_df(m.brm)
Bayes_mean <- round(mean(m.brm_samples$b_Intercept), 2)

hdi <- hdi(m.brm_samples$b_Intercept, credMass = 0.95)
Bayes_lb <- round(as.numeric(hdi[1]), 2)
Bayes_ub <- round(as.numeric(hdi[2]), 2)
```

## Abstract
### Understanding uncertainty: Comparative Bayesian and frequentist meta-analysis of epidemiological studies on lead exposure and IQ loss in children

### Objective
Evidence synthesis in epidemiology is often done by meta-analysis, primarily using a frequentist approach. 
Bayesian meta-analysis provides a framework for uncertainty quantification that incorporates both within-study and between-study variability and results in posterior distributions, facilitating transparent communication of uncertainty. 
The aim of this study is to compare a frequentist to a Bayesian approach and investigate how these frameworks influence exposure-response estimates, uncertainty characterization, and applicability to further use of the results. 
The use case is the association of blood lead levels and children’s neurodevelopment.

### Material and Methods
We conducted a systematic review searching PubMed and Scopus for studies on the association of blood lead levels (BLL) and children’s neurodevelopment (operationalized as IQ score). 
Risk of bias (RoB) was assessed using the rapid RoB tool [@plate2024DevelopmentToolRapid]. 
All regression coefficients were harmonized to the $\log_e$​ scale before pooling.
We assumed a multilevel structure in both models and included the same effect estimates.
For the frequentist approach, we used a random-effects model and parameters were estimated via the restricted maximum-likelihood estimation.
For the Bayesian meta-analysis, we employed a hierarchical Bayesian random-effects model with weakly-informative priors. 
Posterior distributions of parameters were estimated using Markov Chain Monte Carlo (MCMC) sampling.
We conducted additional analyses to examine the sensitivity of the results to (1) possibly biased study results (RoB) and (2) the choice of priors. 

### Results
After title/abstract and full-text screening, we included 12 studies in the meta-analysis.
The frequentist approach lead to a pooled estimate of `r freq_mean` (95% confidence interval (CI): `r freq_lb`; `r freq_ub`) IQ points per unit increase in $\log_e$-transformed BLL (in µg/dL).
The Bayesian model yielded a mean pooled estimate of `r Bayes_mean` (95% high density interval (HDI): `r Bayes_lb`; `r Bayes_ub`). 
Results from the sensitivity analyses did not differ from the main analysis to a relevant extend, suggesting robustness of the results to RoB and the choice of priors in the Bayesian model.
- Heterogeneity 

### Conclusion
While the frequentist approach needed less computational power, the Bayesian approach was less sensitive to outlier studies that reported high effect estimates with wide confidence intervals, resulting in a lower pooled effect compared to the frequentist analysis. (good approach for small number of studies to pool)
The posterior distribution of the Bayesian approach reflects uncertainty directly and thus can be interpreted intuitively in probability terms.
This is especially useful for policy making, because a relevant threshold risk can be defined and the respective probability mass can be estimated. E.g.: comparing the health risk to that of another pollutant.
- Posterior distribution can be interpreted intuitively, in probability terms.
- Lead effect characterized: updated with new studies & posterior distribution can be used e.g. for burden of disease studies. 

Funding
