---
title: "Abstract ISEE 2026"
author:
  - name: Paulina Sell
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0009-0007-7419-2956
  - name: Margaux Sanchez
    affiliation: Agence nationale de sécurité sanitaire, Maisons-Alfort, France
    orcid: 
  - name: Philippe Palmont
    affiliation: Agence nationale de sécurité sanitaire, Maisons-Alfort, France
    orcid: 
  - name: Dietrich Plass
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0000-0002-8058-1399
date-modified: last-modified
date-format: "D MMMM YYYY"
bibliography: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/references.bib
csl: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/vancouver.csl
reference-section-title: "References"
format:
    # html:
    #     toc: false
    #     title-block-banner: false
    #     number-sections: false
    #     embed-resources: true
    #     css: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/custom.scss
    #     page-layout: full
    #     theme: yeti
    docx
execute:
  cache: false
---

## Information for Authors:
* Deadline: February 27, 2026 (23:59 CET)
* Use acronyms only when necessary and define all non-standard abbreviations and concepts in your abstract at first use
* Abstract titles can contain a maximum of **20 words** (no abbreviations)
* The abstract body should not exceed **300 words** and should consist of the four sections: Objective, Material and Methods, Results, Conclusion.
* Phrases such as “results will be discussed” or “data will be presented” should not be used.
* Please mention all sources of funding for the work described and other acknowledgements briefly at the end of the abstract text.

* A **main topic** that is relevant for your abstract must be selected during the submission process. 
The Scientific Organizing Committee reserves the right to change the topics during the abstract review process.
* Please enter **2 to 5 keywords** from the official keyword list

* Topic: A-10: Chemical exposures / A-18: Methods (?)
* Keywords: Chemical exposures, Environmental epidemiology, Heavy metals, Lead, International collaboration, Modeling (?)

```{r}
#| label: read_model
#| echo: false

m.brm <- readRDS("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/models/main/m.brm_full.rds")
freq_model <- readRDS("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/models/main/freq_full.rds")
```

```{r}
#| label: extract_results
#| echo: false
#| message: false


library(posterior)
library(HDInterval)
library(brms)

data = read.csv("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/data/study_data_leadIQloss.csv")

freq_mean <- round(as.numeric(freq_model$beta), 2)
freq_lb <- round(as.numeric(freq_model$ci.lb), 2)
freq_ub <- round(as.numeric(freq_model$ci.ub), 2)

m.brm_samples <- as_draws_df(m.brm)
Bayes_mean <- round(mean(m.brm_samples$b_Intercept), 2)

hdi <- hdi(m.brm_samples$b_Intercept, credMass = 0.95)
Bayes_lb <- round(as.numeric(hdi[1]), 2)
Bayes_ub <- round(as.numeric(hdi[2]), 2)

# calculate tau2 and I2 analogous to metafor
# Extract posterior samples of tau
tau_samples <- as_draws_df(m.brm, variable = "sd_author_year__Intercept")
tau_sq_samples <- tau_samples$sd_author_year__Intercept^2

# Get the SAME typical within-study variance that metafor used
vt <- freq_model$vt

# Calculate I² using metafor's approach for each posterior draw
I2_samples <- tau_sq_samples / (tau_sq_samples + vt)

# Summarize tau²
tau_sq_mean <- mean(tau_sq_samples)
tau_sq_median <- median(tau_sq_samples)
tau_sq_ci <- quantile(tau_sq_samples, c(0.025, 0.975))

# Summarize I²
I2_posterior_mean <- mean(I2_samples)
I2_posterior_median <- median(I2_samples)
I2_credible_interval <- quantile(I2_samples, c(0.025, 0.975))


```

## Abstract
### Quantifying uncertainty: A methodological comparison of Bayesian and frequentist meta-analyses of lead exposure and children's neurodevelopment

### Objective
<!-- Meta-analyses are critical for evidence synthesis in epidemiology. 
While they are primarily conducted in a frequentist framework, Bayesian meta-analysis offers improved uncertainty quantification, incorporating within-study and between-study variability and facilitating transparent communication of uncertainty.
The aim of this study is to compare frequentist and Bayesian meta-analysis frameworks and investigate how they differ in exposure-response estimates, uncertainty characterization, and applicability for risk assessment.
We used the association of blood lead levels and children's neurodevelopment as a use-case. -->

Bayesian meta-analysis provides probabilistic uncertainty quantification valuable for risk assessment and policy-making.
Yet, frequentist approaches dominate in environmental epidemiology. 
We compared frequentist and Bayesian frameworks to evaluate differences in pooled effect estimates, uncertainty characterization, and applicability for risk assessment, using blood lead exposure and children's neurodevelopment as a case study.

### Material and Methods
We systematically searched PubMed and Scopus for epidemiological studies on lead exposure and IQ score in children. 
Risk of bias (RoB) was assessed and regression coefficients were harmonized to the $\log_e$​ scale.
In both meta-analyses, we assumed a multilevel structure, used random-effects models and included identical effect estimates.
The frequentist approach used restricted maximum-likelihood estimation.
The Bayesian meta-analysis employed a hierarchical model with weakly-informative priors.
Sensitivity analyses examined robustness to RoB and prior specification.
$I^2$ was estimated as a measure of between-study heterogeneity.

### Results
Twelve studies were included after screening.
The frequentist approach yielded a pooled estimate of `r freq_mean` (95% confidence interval (CI): `r freq_lb`; `r freq_ub`) IQ points per unit increase in $\log_e$-transformed BLL (µg/dL).
The Bayesian model yielded a mean pooled estimate of `r Bayes_mean` (95% credible interval (CrI): `r Bayes_lb`; `r Bayes_ub`).
Between-study heterogeneity was substantial (frequentist: $\tau^2$ = 3.79, $\text{I}^2$ = 98.5%; Bayesian: $\tau^2$ = `r round(tau_sq_mean, 2)` [95% CrI: `r round(tau_sq_ci[1], 2)`; `r round(tau_sq_ci[2], 2)`], $\text{I}^2$ = `r round(I2_posterior_mean, 3) * 100`% [95% CrI: `r round(I2_credible_interval[1], 3) * 100`%; `r round(I2_credible_interval[2], 3) * 100`%]).
The Bayesian approach provided a posterior distribution for heterogeneity, quantifying uncertainty from between-study variation. 
Sensitivity analyses showed robustness to RoB and prior specification.
The Bayesian approach was less sensitive to outlier studies, resulting in a lower pooled effect.

### Conclusion
In this study, we provide updated pooled estimates for lead's effect on children's neurodevelopment.  
The Bayesian approach enables direct probability statements through posterior distributions.
For risk assessment and policy making, this allows defining risk thresholds and calculating exceedance probabilities, enabling comparative assessments across pollutants.

Funded under EU Horizon (Grant Agreement No 101057014).


<!-- [95% CrI: `r round(tau_sq_ci[1], 2)`; `r round(tau_sq_ci[2], 2)`]
[95% CrI: `r round(I2_credible_interval[1], 3) * 100`%; `r round(I2_credible_interval[2], 3) * 100`%] -->
