---
title: "Abstract ISEE 2026"
author:
  - name: Paulina Sell
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0009-0007-7419-2956
  - name: Margaux Sanchez
    affiliation: Agence nationale de sécurité sanitaire, Maisons-Alfort, France
    orcid: 
  - name: Philippe Palmont
    affiliation: Agence nationale de sécurité sanitaire, Maisons-Alfort, France
    orcid: 
  - name: Dietrich Plass
    affiliation: Environmental Hygiene, German Environment Agency, Berlin, Germany
    orcid: 0000-0002-8058-1399
date-modified: last-modified
date-format: "D MMMM YYYY"
bibliography: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/references.bib
csl: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/vancouver.csl
reference-section-title: "References"
format:
    # html:
    #     toc: true
    #     title-block-banner: false
    #     number-sections: false
    #     embed-resources: true
    #     css: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/custom.scss
    #     page-layout: full
    #     theme: yeti
    docx
execute:
  cache: false
---

## Information for Authors:
* Deadline: February 27, 2026 (23:59 CET)
* The submitting author is required to ensure that all co-authors are aware of the content of the abstract before submission
* Use acronyms only when necessary and define all non-standard abbreviations and concepts in your abstract at first use
* Abstract titles can contain a maximum of **20 words** (no abbreviations)
* The abstract body should not exceed **300 words** and should consist of the four sections: Objective, Material and Methods, Results, Conclusion.
* Phrases such as “results will be discussed” or “data will be presented” should not be used.
* Please mention all sources of funding for the work described and other acknowledgements briefly at the end of the abstract text.

* A main topic that is relevant for your abstract must be selected during the submission process. 
The Scientific Organizing Committee reserves the right to change the topics during the abstract review process.
* Please enter at least 2 to 5 keywords from the official keyword list

* Topic: A-10: Chemical exposures / A-18: Methods ?
* Keywords: Chemical exposures, Environmental epidemiology, Heavy metals, Lead, International collaboration, Modeling

```{r}
#| label: read_model
#| echo: false

m.brm <- readRDS("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/models/main/m.brm_full.rds")
freq_model <- readRDS("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/models/main/freq_full.rds")
```

```{r}
#| label: extract_results
#| echo: false
#| message: false


library(posterior)
library(HDInterval)
library(brms)

data = read.csv("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/data/study_data_leadIQloss.csv")

freq_mean <- round(as.numeric(freq_model$beta), 2)
freq_lb <- round(as.numeric(freq_model$ci.lb), 2)
freq_ub <- round(as.numeric(freq_model$ci.ub), 2)

m.brm_samples <- as_draws_df(m.brm)
Bayes_mean <- round(mean(m.brm_samples$b_Intercept), 2)

hdi <- hdi(m.brm_samples$b_Intercept, credMass = 0.95)
Bayes_lb <- round(as.numeric(hdi[1]), 2)
Bayes_ub <- round(as.numeric(hdi[2]), 2)

# Get tau-squared from tau
tau_samples <- as_draws_df(m.brm, variable = "sd_author_year__Intercept")
tau_sq_samples <- tau_samples$sd_author_year__Intercept^2

# Posterior summary of tau-squared
tau_sq_mean <- mean(tau_sq_samples) # Compare to 3.7946
tau_sq_ci <- quantile(tau_sq_samples, c(0.025, 0.975)) # 95% CrI

# Calculate I² for each study and each posterior draw
n_studies <- nrow(data)
n_samples <- nrow(tau_samples) 

I2_by_study <- matrix(NA, nrow = n_samples, ncol = n_studies)

data$vi <- data$se_beta_ln^2

for(i in 1:n_studies) {
  I2_by_study[, i] <- tau_sq_samples / (tau_sq_samples + data$vi[i])
}

# Average I² across studies for each posterior draw
I2_samples <- rowMeans(I2_by_study)

# Summarize the posterior distribution
I2_posterior_mean <- mean(I2_samples)
I2_posterior_median <- median(I2_samples)
I2_credible_interval <- quantile(I2_samples, c(0.025, 0.975))

```

## Abstract
### Understanding uncertainty: Comparative Bayesian and frequentist meta-analysis of epidemiological studies on lead exposure and IQ loss in children

### Objective
Evidence synthesis in epidemiology is often done by meta-analysis, primarily using a frequentist approach. 
Bayesian meta-analysis offers an alternative framework for uncertainty quantification, incorporating within-study and between-study variability and facilitating transparent communication of uncertainty.
We compared frequentist and Bayesian frameworks to investigate how they differ in exposure-response estimates, uncertainty characterization, and applicability for risk assessment.
We used the association of blood lead levels and children's neurodevelopment as a use-case.

### Material and Methods
We systematically searched PubMed and Scopus for epidemiological studies on blood lead levels and IQ score in children. 
Risk of bias (RoB) was assessed using a rapid RoB tool [@plate2024DevelopmentToolRapid] and regression coefficients were harmonized to the $\log_e$​ scale.
We assumed a multilevel structure and used random-effects models.
We included the same effect estimates for both approaches.
The frequentist approach used restricted maximum-likelihood estimation.
The Bayesian meta-analysis employed a hierarchical model with weakly-informative priors. 
Posterior distributions of parameters were obtained using Markov Chain Monte Carlo.
Sensitivity analyses examined robustness to RoB and prior specification.
$I^2$ was calculated as a measure of between-study heterogeneity in the frequentist approach and directly modelled in the Bayesian approach.

### Results
Twelve studies were included after screening.
The frequentist approach yielded a pooled estimate of `r freq_mean` (95% confidence interval (CI): `r freq_lb`; `r freq_ub`) IQ points per unit increase in $\log_e$-transformed BLL (in µg/dL).
The Bayesian model yielded a mean pooled estimate of `r Bayes_mean` (95% high density interval (HDI): `r Bayes_lb`; `r Bayes_ub`).
Between-study heterogeneity was substantial (frequentist: $\tau^2$ = 3.79, $\text{I}^2$ = 98.5%; Bayesian: $\tau^2$ = `r round(tau_sq_mean, 2)` [95% credible interval (CrI): `r round(tau_sq_ci[1], 2)`; `r round(tau_sq_ci[2], 2)`], $\text{I}^2$ = `r round(I2_posterior_mean, 3) * 100`% [95% CrI: `r round(I2_credible_interval[1], 3) * 100`%; `r round(I2_credible_interval[2], 3) * 100`%]).
The Bayesian approach provided a posterior distribution for heterogeneity, quantifying uncertainty from between-study variation. 
Sensitivity analyses showed robustness to RoB assessment and prior choice.
The Bayesian approach was less sensitive to outlier studies that reported wide confidence intervals, resulting in a lower pooled effect compared to the frequentist analysis.

### Conclusion
In this study, we provide updated pooled estimates for lead's effect on children's neurodevelopment.  
The Bayesian approach offered distinct advantages for uncertainty quantification as the posterior distributions enable direct probability statements.
This is especially useful for risk assessments and policy making, as it allows defining relevant risk thresholds and calculating the probability of exceeding them, enabling comparative risk evaluation across pollutants.

Funding
