---
title: "Abstract ISEE 2026"
author:
  - name: Paulina Sell
    affiliations: 
      - ref: uba
    email: paulina.sell@uba.de
    orcid: 0009-0007-7419-2956
  - name: Margaux Sanchez
    affiliations: 
      - ref: anses
    orcid: 
  - name: Philippe Palmont
    affiliations: 
      - ref: anses
    orcid: 
  - name: Dietrich Plass
    affiliations:
      - ref: uba
    orcid: 0000-0002-8058-1399

affiliations:
  - id: uba
    name: German Environment Agency, Environmental Hygiene, Berlin, Germany
  - id: anses
    name: French National Agency for Food, Environmental and Occupational Health & Safety, Maisons-Alfort, France

date-modified: last-modified
date-format: "D MMMM YYYY"
bibliography: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/references.bib
csl: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/vancouver.csl
reference-section-title: "References"

filters:
 - authors-block

format:
    # html:
    #     toc: false
    #     title-block-banner: false
    #     number-sections: false
    #     embed-resources: true
    #     css: /Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/manuscript/custom.scss
    #     page-layout: full
    #     theme: yeti
    docx: default
execute:
  cache: false
---

## Information for Authors:
* Deadline: February 27, 2026 (23:59 CET)
* Use acronyms only when necessary and define all non-standard abbreviations and concepts in your abstract at first use
* Abstract titles can contain a maximum of **20 words** (no abbreviations)
* The abstract body should not exceed **300 words** and should consist of the four sections: Objective, Material and Methods, Results, Conclusion.
* Phrases such as “results will be discussed” or “data will be presented” should not be used.
* Please mention all sources of funding for the work described and other acknowledgements briefly at the end of the abstract text.

* A **main topic** that is relevant for your abstract must be selected during the submission process. 
The Scientific Organizing Committee reserves the right to change the topics during the abstract review process.
* Please enter **2 to 5 keywords** from the official keyword list

* Topic: A-10: Chemical exposures / A-18: Methods (?)
* Keywords: Chemical exposures, Environmental epidemiology, Heavy metals, Lead, International collaboration, Modeling (?)

```{r}
#| label: read_model
#| echo: false

m.brm <- readRDS("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/models/main/m.brm_full.rds")
freq_model <- readRDS("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/models/main/freq_full.rds")
```

```{r}
#| label: extract_results
#| echo: false
#| message: false


library(posterior)
library(HDInterval)
library(brms)
library(metafor)

data = read.csv("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/data/study_data_leadIQloss.csv")

freq_mean <- round(as.numeric(freq_model$beta), 2)
freq_lb <- round(as.numeric(freq_model$ci.lb), 2)
freq_ub <- round(as.numeric(freq_model$ci.ub), 2)

# tau2 for frequentist model
freq_confint <- confint(freq_model)
freq_confint <- freq_confint$random

m.brm_samples <- as_draws_df(m.brm)
Bayes_mean <- round(mean(m.brm_samples$b_Intercept), 2)

hdi <- hdi(m.brm_samples$b_Intercept, credMass = 0.95)
Bayes_lb <- round(as.numeric(hdi[1]), 2)
Bayes_ub <- round(as.numeric(hdi[2]), 2)

# calculate tau2 and I2 analogous to metafor
# Extract posterior samples of tau
tau_samples <- as_draws_df(m.brm, variable = "sd_author_year__Intercept")
tau_sq_samples <- tau_samples$sd_author_year__Intercept^2

# Get the SAME typical within-study variance that metafor used
vt <- freq_model$vt

# Calculate I² using metafor's approach for each posterior draw
I2_samples <- tau_sq_samples / (tau_sq_samples + vt)

# Summarize tau²
tau_sq_mean <- mean(tau_sq_samples)
tau_sq_median <- median(tau_sq_samples)
tau_sq_ci <- quantile(tau_sq_samples, c(0.025, 0.975))

# Summarize I²
I2_posterior_mean <- mean(I2_samples)
I2_posterior_median <- median(I2_samples)
I2_credible_interval <- quantile(I2_samples, c(0.025, 0.975))
```

## Abstract
### Quantifying uncertainty: A methodological comparison of Bayesian and frequentist meta-analyses of studies on lead exposure and children's intelligence quotient

### Objective
Quantifying uncertainty is crucial for risk assessment and policy-making but current meta-analytic approaches fall short in this regard.
Bayesian meta-analysis provides probabilistic uncertainty quantification.
To evaluate differences in uncertainty quantification and applicability for risk assessment, we compared frequentist and Bayesian frameworks, using lead exposure and children's intelligence quotient (IQ) as a case study.

### Material and Methods
We systematically searched PubMed and Scopus for epidemiological studies on blood lead level (BLL) and IQ score in children. 
Risk of bias (RoB) was assessed and regression coefficients were harmonized to the $\log_e$​ scale.
In both meta-analyses, we assumed a multilevel structure, used random-effects models and included the same effect estimates.
The frequentist approach used restricted maximum-likelihood estimation.
The Bayesian meta-analysis employed a hierarchical model with weakly-informative priors.
Sensitivity analyses examined robustness to RoB and prior specification.
$\tau^2$ and $I^2$ were estimated as measures of between-study heterogeneity.

### Results
Twelve studies were included after screening.
The frequentist approach yielded a pooled estimate of `r freq_mean` (95% confidence interval (CI): `r freq_lb`; `r freq_ub`) IQ points per unit increase in $\log_e$-transformed BLL (µg/dL).
The Bayesian model yielded a mean pooled estimate of `r Bayes_mean` (95% credible interval (CrI): `r Bayes_lb`; `r Bayes_ub`).
Between-study heterogeneity was substantial (frequentist: $\tau^2$ = `r round(freq_confint[1,1], 1)`, $\text{I}^2$ = `r round(freq_confint[3,1], 1)`%; Bayesian: $\tau^2$ = `r round(tau_sq_mean, 2)`, $\text{I}^2$ = `r round(I2_posterior_mean, 3) * 100`%).
The Bayesian approach provided a posterior distribution for heterogeneity, quantifying uncertainty from between-study variation. 
Sensitivity analyses showed robustness to RoB and prior specification.
The Bayesian approach was less sensitive to outlier studies, resulting in a lower pooled effect.

### Conclusion
We provide updated pooled estimates for the effect of lead exposure on children's IQ.
The Bayesian approach provides posterior distributions, enabling direct probability statements, for example about exceeding risk thresholds, and comparative assessments across pollutants.
These features are of particular value regarding risk assessment and policy-making.

Funded under EU Horizon (Grant Agreement No 101057014).


<!-- [95% CrI: `r round(I2_credible_interval[1], 3) * 100`%; `r round(I2_credible_interval[2], 3) * 100`%] -->
<!-- [95% CrI: `r round(tau_sq_ci[1], 2)`; `r round(tau_sq_ci[2], 2)`] -->
