---
title: "Abstract StanCon 2026"
author:
  - name: Paulina Sell
    affiliations: 
      - ref: uba
    corresponding: true
    email: paulina.sell@uba.de
    adress: Corrensplatz 1, 14195 Berlin, Germany
    orcid: 0009-0007-7419-2956
  - name: Margaux Sanchez
    affiliations: 
      - ref: anses
    orcid: 0000-0002-1923-6393
  - name: Philippe Palmont
    affiliations: 
      - ref: anses
    orcid: 0009-0005-5624-3460
  - name: Simon Steiger
    affiliations:
      - ref: ki
    orcid: 0009-0003-9745-354X
  - name: Dietrich Plass
    affiliations:
      - ref: uba
    orcid: 0000-0002-8058-1399

affiliations:
  - id: uba
    name: German Environment Agency, Environmental Hygiene, Berlin, Germany
  - id: anses
    name: French National Agency for Food, Environmental and Occupational Health & Safety, Maisons-Alfort, France
  - id: ki
    name: Karolinska Institutet, Department of Medicine Solna, Division of Clinical Epidemiology, Stockholm, Sweden

date-modified: last-modified
date-format: "D MMMM YYYY"

filters:
 - authors-block

format:
    # html:
    #     toc: false
    #     title-block-banner: false
    #     number-sections: false
    #     embed-resources: true
    #     css: ../manuscript/custom.scss
    #     page-layout: full
    #     theme: yeti
    docx: default
execute:
  cache: false
---

[Link to conference](https://www.stancon2026.org)

## Information for Authors

* Title: The abstract title is limited to **150 characters**. Do not use capital letters unless abbreviations or similar.
* Authors and affiliations: Please enter authors’ name and affiliations in the dedicated fields. Do not use capital letters unless abbreviations or similar.
* Text: Word count is limited to **350 words**. The title, authors’ name and affiliation are not included in this number.
* Image: The abstract allows for no images/tables/figures.
 
## Poster Presentation
Poster presentations provide an informal setting to share work, spark conversations, and get detailed feedback. Participants can engage directly with authors during dedicated poster sessions. This format encourages in-depth technical exchange at an approachable pace.

```{r}
#| label: read_model
#| echo: false

m.brm <- readRDS("../models/main/m.brm_full.rds")
freq_model <- readRDS("../models/main/freq_full.rds")
```

```{r}
#| label: extract_results
#| echo: false
#| message: false


library(posterior)
library(HDInterval)
library(brms)
library(metafor)

data = read.csv("../data/study_data_leadIQloss.csv")

freq_mean <- round(as.numeric(freq_model$beta), 2)
freq_lb <- round(as.numeric(freq_model$ci.lb), 2)
freq_ub <- round(as.numeric(freq_model$ci.ub), 2)

# tau2 for frequentist model
freq_confint <- confint(freq_model)
freq_confint <- freq_confint$random

m.brm_samples <- as_draws_df(m.brm)
Bayes_mean <- round(mean(m.brm_samples$b_Intercept), 2)

hdi <- hdi(m.brm_samples$b_Intercept, credMass = 0.95)
Bayes_lb <- round(as.numeric(hdi[1]), 2)
Bayes_ub <- round(as.numeric(hdi[2]), 2)

# calculate tau2 and I2 analogous to metafor
# Extract posterior samples of tau
tau_samples <- as_draws_df(m.brm, variable = "sd_author_year__Intercept")
tau_sq_samples <- tau_samples$sd_author_year__Intercept^2

# Get the SAME typical within-study variance that metafor used
vt <- freq_model$vt

# Calculate I² using metafor's approach for each posterior draw
I2_samples <- tau_sq_samples / (tau_sq_samples + vt)

# Summarize tau²
tau_sq_mean <- mean(tau_sq_samples)
tau_sq_median <- median(tau_sq_samples)
tau_sq_ci <- quantile(tau_sq_samples, c(0.025, 0.975))

# Summarize I²
I2_posterior_mean <- mean(I2_samples)
I2_posterior_median <- median(I2_samples)
I2_credible_interval <- quantile(I2_samples, c(0.025, 0.975))
```

\newpage

## Abstract
**Quantifying uncertainty: a methodological comparison of Bayesian and frequentist meta-analyses of studies on lead exposure and children's IQ**

**Objective**:
Uncertainty quantification is central to evidence synthesis for risk assessment and policy-making, but frequentist meta-analytic approaches - currently predominant in epidemiology - often fall short in this regard.
Bayesian meta-analysis provides probabilistic uncertainty quantification.
To evaluate differences in uncertainty quantification and applicability for risk assessment, we compared frequentist and Bayesian frameworks, using lead exposure and children's IQ as a case study.

**Methods**:
We systematically searched PubMed and Scopus for epidemiological studies on blood lead level (BLL) and IQ score in children. 
After screening, risk of bias (RoB) in the identified studies was assessed and regression coefficients were harmonized to the $\log_e$​ scale.
In both meta-analyses, we included the same effect estimates, assumed a multilevel structure and used random-effects models.
In the frequentist approach, we used the R package `metafor` and restricted maximum-likelihood estimation.
The Bayesian meta-analysis employed a hierarchical model implemented in `brms`, an R interface to `Stan`, with weakly-informative priors.
Sensitivity analyses examined robustness to RoB and to prior specification.
$\tau^2$ and $I^2$ were estimated as measures of between-study heterogeneity.

**Results**:
Twelve studies were included in the meta-analysis.
The frequentist approach yielded a pooled estimate of `r freq_mean` (95% confidence interval (CI): `r freq_lb`; `r freq_ub`) IQ points per unit increase in $\log_e$-transformed BLL (µg/dL).
The Bayesian model yielded a mean pooled estimate of `r Bayes_mean` (95% credible interval (CrI): `r Bayes_lb`; `r Bayes_ub`).
Between-study heterogeneity was substantial (frequentist: $\tau^2$ = `r round(freq_confint[1,1], 1)`, $\text{I}^2$ = `r round(freq_confint[3,1], 1)`%; Bayesian: $\tau^2$ = `r round(tau_sq_mean, 2)`, $\text{I}^2$ = `r round(I2_posterior_mean, 3) * 100`%).
The Bayesian approach provided a full posterior distribution for heterogeneity, directly quantifying uncertainty from between-study variation. 
Sensitivity analyses showed robustness to RoB and prior specification.
The Bayesian approach was less sensitive to outlier studies reporting substantial uncertainty, resulting in a lower pooled effect compared to the frequentist approach.
This regularisation is particularly valuable in small meta-analyses, where frequentist heterogeneity estimates are unstable — though it relies on heterogeneity priors that are carefully grounded in domain knowledge.

**Conclusion**:
We provide updated pooled estimates for the effect of lead exposure on children's IQ.
The Bayesian approach provides posterior distributions, enabling direct probability statements, for example about exceeding risk thresholds, and comparative assessments across pollutants.
These features are of particular value regarding risk assessment and policy-making.