---
title: "Bayesian Meta Analysis on Lead Exposure & IQ Loss"
author: "Paulina Sell"
# date: "2025-07-25"
date-modified: "2025-08-01"
format:
  html:
    toc: true
    self-contained: true
    embed-resources: true
    theme: cosmo
    page-layout: full
execute: 
  cache: refresh
fig-align: center
fig-height: 4
---

```{r}
#| label: loading-data
#| echo: false

data_clean <- read.csv("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/data/study_data_leadIQloss_clean.csv")
```

# Introduction
The aim of this project is to estimate the environmental burden of disease (EBD) attributable to lead exposure in Germany. Our focus will be on the neurodevelopmental effects of lead; therefore, the health outcome will be IQ point loss in children and adolescents. In order to estimate these population health effects, we must find solid evidence for exactly how lead exposure is related to IQ loss. The exposure data we are working with is based on whole blood samples from 720 children and adolescents (ages 3 - 17) in Germany, collected as part of the German Environmental Survey ([GerES V](https://www.umweltbundesamt.de/en/topics/health/assessing-environmentally-related-health-risks/german-environmental-surveys/german-environmental-survey-2014-2017-geres-v#undefined)) between 2014 and 2017.\
\
In order to quantify exactly how lead exposure is related to IQ loss, we turn to the epidemiological literature. Many studies have been published focusing on the neurodevelopmental effects of lead. Rather than selecting an estimate from a single study for our EBD analysis, we will pool estimates from multiple studies in order to gather and strengthen the evidence we base our analysis on. And since uncertainty plays an important role in the final burden of disease results, we will use a Bayesian approach. This approach allows us to easily model uncertainty and obtain a full posterior distribution of the pooled effect (rather than a mean and a hard-to-interpret confidence interval). Later, we will be able to directly use the posterior distribution in our EBD analysis! But first, let's meta-analyze some epi studies.\
I found help and inspiration in multiple sources and have (hopefully) linked them all. I especially found this hands-on guide from [Harrer et al. 2021](https://bookdown.org/MathiasHarrer/Doing_Meta_Analysis_in_R/bayesian-ma.html#bayesian-ma) very helpful.


# Step 1: Standardizing Effect Estimates

We first need to search for and identify relevant studies and extract the data necessary for our analysis. This was already done by Philippe Palmont and his team at ANSES.\
Next, we need to harmonize / standardize the effect estimates before we can pool them. Most of the studies transformed their blood lead level (BLL) data to a logarithmic scale in order to fulfil the assumption of a linear relationship between BLL (exposure) and IQ points (outcome) in their regression analyses. Unfortunately, they did not always use the same log base. Some used log~10~, some ln, and some used log~2~. Also, 3 of the 13 included studies did not transform their BLL at all, so it's on a linear scale. In their analysis, Philippe Palmont and his team standardized all estimates to linear increments of BLL, using the template proposed by [Rodriguez Barranco 2017](https://doi.org/10.1186/s12874-017-0322-8). What happens to the effect estimates, if we transform them in this way?\

## Visually Checking Transformations

Let's compare an original beta coefficient from [Crump et al. 2013](https://doi.org/10.3109/10408444.2013.832726) (ß = -2.65) for ln(BLL) to the transformed one (ß = -0.277), which can now be used for linear BLL.

```{r}
#| label: lin-vs-log
#| echo: false # Don't include source code in output

library(ggplot2)

beta_log <- -2.650
beta_linear <- -0.277

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_log <- beta_log * log(blood_lead + 1)
IQ_loss_linear <- beta_linear * blood_lead

plot_data <- data.frame(
  blood_lead = blood_lead,
  log_ERF = IQ_loss_log,
  linear_ERF = IQ_loss_linear
)

ggplot(plot_data, aes(x = blood_lead)) +
  geom_line(aes(y = log_ERF, color = "Logarithmic")) +
  geom_line(aes(y = linear_ERF, color = "Linear")) +
  scale_color_manual(values = c("Logarithmic" = "blue", "Linear" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

It appears that we would heavily underestimate the effects of lower BLL and overestimate the effects of higher BLL. Since today, BLL are often found to be much lower than even 30 µg/dL (in our sample, the range is approximately 0.04 to 4.5 µg/dL), we would likely underestimate the population health effects of BLL on IQ in children.\
Therefore, transforming log estimates to linear seems to be problematic in our case. 

## Transforming Estimates to ln(BLL)
In order to preserve the correct exposure-response curves and harmonize the effect estimates for the meta analysis, we could try to transform all estimates to ln(BLL), as this is closer to the original estimates reported by the studies, which mostly just used different bases for their logarithmic transformations.   
   
These are the formulas I used to transform the different estimates to ln(BLL):\
$$
\beta_{ln} = \beta_{linear} \times C_{ref}\
$$
$$
\beta_{ln} = \frac{beta_{log_{2}}} {\ln(2)}\
$$
$$
\beta_{ln} = \frac{beta_{log_{10}}} {\ln(10)}\
$$
Where:\
$\beta_{ln}$ = harmonized beta coefficient (natural logarithm)\
$\beta_{linear}$ = beta coefficient for linear BLL\
$\beta_{log_2}$ = beta coefficient for log₂-transformed BLL\
$\beta_{log_{10}}$ = beta coefficient for log₁₀-transformed BLL\
$C_{ref}$ = reference concentration for linear transformation\


### Case 1: log~10~(BLL) -> ln(BLL): [Dantzer et al. 2020](https://doi.org/10.1016/j.scitotenv.2019.134519)
```{r}
#| label: log10-to-ln
#| echo: false


Dantzerbeta_log10 <- -13.150
Dantzerbeta_ln <- -5.711

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_log10 <- Dantzerbeta_log10 * log10(blood_lead + 1)
IQ_loss_ln <- Dantzerbeta_ln * log(blood_lead + 1)

Dantzerplot_data <- data.frame(
  blood_lead = blood_lead,
  log_ERF = IQ_loss_log10,
  ln_ERF = IQ_loss_ln
)

ggplot(Dantzerplot_data, aes(x = blood_lead)) +
  geom_line(aes(y = log_ERF, color = "log(10)")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("log(10)" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

We can't even see the blue curve, so the transformation should be valid.   


### Case 2: log~2~(BLL) -> ln(BLL): [Desrochers-Couture et al. 2018](https://doi.org/10.1016/j.envint.2018.10.043)
```{r}
#| label: log2-to-ln
#| echo: false



Desrochersbeta_log2 <- 0.014
Desrochersbeta_ln <- 0.020

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_log2 <- Desrochersbeta_log2 * log2(blood_lead + 1)
IQ_loss_ln <- Desrochersbeta_ln * log(blood_lead + 1)

Desrochersplot_data <- data.frame(
  blood_lead = blood_lead,
  log_ERF = IQ_loss_log2,
  ln_ERF = IQ_loss_ln
)

ggplot(Desrochersplot_data, aes(x = blood_lead)) +
  geom_line(aes(y = log_ERF, color = "log(2)")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("log(2)" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

For the transformation from log~2~ to ln, we can see that at least in this concentration range, the transformed estimate is a good approximation of the original.\
**!** Note that this study reported a small protective effect of lead exposure, this is why the curve is flipped.   

### Case 3: linear(BLL) -> ln(BLL): [Iglesias et al. 2011](https://doi.org/10.1179/107735211799041841)
In order to transform this estimate, we need to set a reference BLL ($C_{ref}$). Here, I chose 2.2 µg/dL.  
```{r}
#| label: lin-to-ln
#| echo: false


Iglesiasbeta_linear <- -0.940
Iglesiasbeta_ln <- -2.068

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_linear <- Iglesiasbeta_linear * blood_lead
IQ_loss_ln <- Iglesiasbeta_ln * log(blood_lead + 1)

Iglesiasplot_data <- data.frame(
  blood_lead = blood_lead,
  linear_ERF = IQ_loss_linear,
  ln_ERF = IQ_loss_ln
)

ggplot(Iglesiasplot_data, aes(x = blood_lead)) +
  geom_line(aes(y = linear_ERF, color = "linear")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("linear" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

This looks problematic... However, the concentration range is quite high in this plot, so let's zoom in.

```{r}
#| label: lin-to-ln-zoom
#| echo: false


blood_lead_low <- seq(0, 5, by = 0.1)

IQ_loss_linear <- Iglesiasbeta_linear * blood_lead_low
IQ_loss_ln <- Iglesiasbeta_ln * log(blood_lead_low + 1)

Iglesiasplot_data <- data.frame(
  blood_lead_low = blood_lead_low,
  linear_ERF = IQ_loss_linear,
  ln_ERF = IQ_loss_ln
)

ggplot(Iglesiasplot_data, aes(x = blood_lead_low)) +
  geom_line(aes(y = linear_ERF, color = "linear")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("linear" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

Even in the lower concentration range, it appears that the transformed estimate does not accurately approximate the original estimate. For my meta analysis, I therefore decided to exclude 3/13 studies with estimates for untransformed BLL data.\
This leaves us with 10 harmonized effect estimates for the association of ln(BLL) and absolute IQ point loss.\
\
These are the final studies and transformed effect estimates that we will be working with.
```{r}
#| label: view-data
#| echo: false

knitr::kable(data_clean, digits = 2)
```


# Step 2: Meta Analysis
Now that we have collected our study data and harmonized the effect estimates, we can finally begin our meta analysis!
```{r}
#| label: load-packages-data
#| echo: false
#| message: false


# loading required packages
library(brms)
library(tidyverse)
library(tidybayes)
library(ggridges)
library(glue)
library(extraDistr)
library(bayesplot)
library(HDInterval)

# loading data
data <- read.csv("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/data/study_data_leadIQloss.csv")

# try excluding studies that were transformed from linear to log, since transformation may not be valid here (and Roy, because I don't trust the estimate.. asked main author for confirmation / correction)
data <- data[!data$author_year %in% c("Halabicky 2022", "Iglesias 2011", "Min 2009", "Roy 2013"), ]

```

```{r}
#| label: def-values
#| echo: false

mean <- 0
sd <- 1

lower <- -2
upper <- 2
```

## Setting Priors
First, we need to define the priors. In Bayesian statistics, a prior distribution represents our initial belief about an unknown parameter before observing any data. It's a probability distribution reflecting our knowledge or assumptions about the parameter's possible values. This prior distribution is then combined with the likelihood function (based on observed data) to produce a posterior distribution, which represents our updated belief after considering the data. You can read more about choosing prior distributions in Bayesian statistics in this [book chapter](https://bookdown.org/kevin_davisross/bayesian-reasoning-and-methods/prior.html#prior).\
\
We will be conservative and use weakly informative priors. For the main effect ($μ$), I chose $μ \sim N(0, 1)$. This means that we assume the main effect to lie between `r lower` and `r upper`, with a `r round(pnorm(upper, mean, sd) - pnorm(lower, mean, sd), 3) * 100`% probability.

```{r}
#| label: vis-main-effect
#| echo: false


# main effect, beta
# visualize distr
x <- seq(-3, 3, by = 0.1)
y <- dnorm(x, mean, sd)
ggplot() +
  aes(x, y) +
  geom_line(colour = "blue") +
  theme_minimal() +
  labs(x= NULL, y = NULL)
```

Setting the prior for heterogeneity ($τ$) is slightly more challenging since, by definition, it must be positive. Here, a half-Cauchy distribution can be useful. It is defined by two parameters: the location, which determines the center of the distribution, and the scale parameter, which controls the spread. We set the lower limit to zero (`lb = 0`) to make it _half_. For this model, I chose $τ \sim HC(0, 0.5)$.

```{r}
#| label: vis-heterogeineity
#| echo: false


# heterogeneity
sigma <- 0.5

# visualize distr
x <- seq(0, 4, by = 0.01)
y <- dhcauchy(x, sigma = sigma)
ggplot() +
  aes(x, y) +
  geom_line(colour = "orange") +
  theme_minimal() +
  labs(x= NULL, y = NULL)
```

```{r}
#| label: phcauchy
#| echo: false

threshold <- 1.5
prob <- (1 - phcauchy(threshold, sigma)) * 100
```
As you can see, we assign a fairly high probability (`r round(prob, 2)`%) to the heterogeneity being higher than `r threshold`. Considering that the median standard error extracted from our included studies is `r round(median(data$se_beta_ln), 2)`, this prior is weakly informative.

This is how we set our priors:

```{r}
#| label: set-priors


priors <- c(prior(normal(0,1), class = Intercept), 
            prior(cauchy(0,0.5), class = sd, lb = 0))
```

## Fitting the model
We can now write and fit the model, using the package _brms_ by [Bürkner 2017](https://doi.org/10.18637/jss.v080.i01). In the second line, we specify our model. We set `adapt_delta` to 0.99 to decrease the step size that the sampling algorithm explores the posterior distribution with. If we don't do this, adapt_delta will default to 0.8 and we get something like `Warning: There were 5 divergent transitions after warmup`. This means that the algorithm is having difficulty exploring the posterior distribution.

```{r}
#| label: fit-main-model
#| output: false

 
m.brm <- brm(
  beta_ln|se(se_beta_ln) ~ 1 + (1|author_year),
  data = data,
  prior = priors,
  control = list(adapt_delta = 0.99),
  save_pars = save_pars(all = TRUE),
  chains = 4,
  iter = 4000,
  seed = 1223)
```

## Assessing Convergence & Model Validity
First, let's make sure that our priors are reasonable. We can do this by doing a prior predictive check. Prior predictive checks are used to evaluate the suitability of prior distributions before the model 'sees' our data. This involves simulating data from the model using only the prior distributions and then examining the simulated data to see if it aligns with our expectations. This can help to identify issues with the chosen priors, such as implausible parameter values, before they can influence the posterior distribution and final inferences. You can read more about prior and posterior predictive checks [here](https://mc-stan.org/docs/stan-users-guide/posterior-predictive-checks.html).

```{R}
#| label: fit-prior-pred
#| output: false
#| echo: false

 
# Fit model again and only sample from priors for prior predictive check
fitPrior <- brm(
  beta_ln|se(se_beta_ln) ~ 1 + (1|author_year), 
  data = data, 
  prior = priors,
  control = list(adapt_delta = 0.99),
  sample_prior = "only",
  chains = 4,
  iter = 4000,
  seed = 1223)
```

```{r}
#| label: prior-pred-plot
#| echo: false


pp_check(fitPrior, ndraws = 20)
```

Now let's do a posterior predictive check:

```{r}
#| label: post-pred-plot
#| echo: false


pp_check(m.brm, ndraws = 20)
```

We mainly want to pay attention to extreme deviations from the area we'd expect our curves to land. Both look okay.\
\
Let's look at the results summary:

```{r}
#| label: summary
#| warning: false
#| error: false
#| echo: false

summary(m.brm)
```
Here, we should pay attention to the $\hat{R}$ (Rhat in output) value (potential scale reduction factor), which indicates whether the Markov chain Monte Carlo (MCMC) sampling process (this is the class of algorithms used to estimate the parameters) has converged. A value of less than 1.05 is generally considered good. [Here](https://mc-stan.org/rstan/reference/Rhat.html) you can read more about convergence and efficiency diagnostics for Markov chains. As we can see, our $\hat{R}$ values are both 1.00 which is good!

Now let's look at the trace plots of the four MCMC chains for both the posterior distribution of beta and the standard deviation. The red vertical lines show where the warm-up ended (samples from the warm-up will not be used further).

```{r}
#| label: plot-chains
#| echo: false
#| fig-height: 6

# extract posterior samples with warm-up
posterior_samples_warm = as_draws_df(m.brm, inc_warmup = T)
names(posterior_samples_warm)[names(posterior_samples_warm) == "b_Intercept"] = "beta"
names(posterior_samples_warm)[names(posterior_samples_warm) == "sd_author_year__Intercept"] = "sd"

mcmc_trace(posterior_samples_warm,
           pars = c("beta", "sd"),
           facet_args = list(ncol = 1)) +
  vline_at(2000, color = "red", linetype = 2, size = 0.5)  # mark end of warm-up
```

## Interpreting the Results

Interesting! Now what do our posterior distributions actually look like? 

```{r}
#| label: plot_post_distr
#| echo: false

# extract posterior samples with warm-up
posterior_samples = as_draws_df(m.brm)
names(posterior_samples)[names(posterior_samples) == "b_Intercept"] = "beta"
names(posterior_samples)[names(posterior_samples) == "sd_author_year__Intercept"] = "sd"

# density plot with HDI
mcmc_areas(posterior_samples,
           pars = c("beta"),
           point_est = ("mean"),
           prob = 0.95) + # 95% HDI
  labs(title = "Posterior Distribution of Beta with Mean & 95% HDI")

mcmc_areas(posterior_samples,
           pars = c("sd"),
           point_est = ("mean"),
           prob = 0.95) + # 95% HDI
  labs(title = "Posterior Distribution of SD with Mean & 95% HDI")
```

The posterior probability distribution represents the updated beliefs about an unknown parameter (e.g. the effect of BLL on IQ, beta) after observing new data, combining prior beliefs with the evidence from the data. And unlike a mean effect + confidence interval, it actually represents probability!\
\
This means that we can 'ask' this distribution many questions, such as: "What is the probability that the effect is greater than -1?" ( _greater_ meaning _stronger_ here, so more IQ points would be lost). To answer such questions, we use the empirical cumulative distribution function (ECDF):

```{r}
#| label: ecdf


b.ecdf <- ecdf(posterior_samples$beta)
x <- b.ecdf(-1)
```

Given our data and model, the probability of the effect being greater than -1 is `r (round(x, 2) * 100)`%.\
\
We can also ask about interesting intervals;  for example, consider the interval we saw in the plot above. This is a 95% high-density interval (HDI), meaning it is the narrowest interval containing 95% of the probability mass. This interval is often referred to as _credible interval_.

```{r}
#| label: HDI
#| output: false

hdi <- hdi(posterior_samples$beta, credMass = 0.95)
```

Given our data and model, the true effect lies between `r round(hdi[1], 2)` and `r round(hdi[2], 2)` with 95% probability.\
\
Coming back to the main objective of this project, we can now use the posterior distribution directly for our EBD assessment!\
\
And of course, we can also plot a pretty forest plot showing all the posterior distributions of the included studies, the mean, and the 95% credible intervals.

```{r}
#| label: forest_plot
#| echo: false
#| warning: false
#| fig-height: 6


# first, we have to prepare the data and extract the posterior distribution for each individual study 
study.draws <- spread_draws(m.brm, r_author_year[author_year,], b_Intercept) %>% 
  mutate(b_Intercept = r_author_year + b_Intercept)

# Generate distribution of the pooled effect
pooled.effect.draws <- spread_draws(m.brm, b_Intercept) %>% 
  mutate(author_year = "Pooled Effect")

# Bind study.draws & pooled.effect.draws to one data frame, start pipe, call ungroup, then use mutate to clean study labels
forest.data <- bind_rows(study.draws, 
                         pooled.effect.draws) %>% 
  ungroup() %>%
  mutate(author_year = str_replace_all(author_year, "[.]", " ")) %>%
  mutate(author_year = fct_rev(factor(author_year)))

# The forest plot should also display the effect size & credible interval of each study. 
# We use forest.data data set, group it by author, and then use the mean_qi function to calculate these values
forest.data.summary <- group_by(forest.data, author_year) %>% 
  mean_qi(b_Intercept)

# let's plot!
ggplot(aes(b_Intercept, 
           relevel(author_year, "Pooled Effect", # Y-axis uses author_year but reorders it so "Pooled Effect" appears at the bottom (after = Inf)
                   after = Inf)), 
       data = forest.data) +

  # Box to highlight pooled effect
  geom_rect(data = filter(forest.data.summary, author_year == "Pooled Effect"),
            aes(xmin = -Inf, xmax = Inf, 
                ymin = as.numeric(factor(author_year)) - 0.4, 
                ymax = as.numeric(factor(author_year)) + 0.4),
            fill = "steelblue3", alpha = 0.2) +
  
  # Add vertical lines for pooled effect and CI
  geom_vline(xintercept = fixef(m.brm)[1, 1], # line at the pooled effect estimate
             color = "gray50", linewidth = 0.8, linetype = 2) + 
  #geom_vline(xintercept = fixef(m.brm)[1, 3:4], # lines for Q2.5 & Q97.5 from pooled effect (95 % CI)
             #color = "gray50", linewidth = 0.8, linetype = 2) + 
  geom_vline(xintercept = 0, color = "gray20", 
             linewidth = 1) + # line at zero (null effect line)
  
  # Add density ridges
  geom_density_ridges(fill = "steelblue3", # ridge density plots for each study/row
                      rel_min_height = 0.01, # removes very small density values
                      col = NA, scale = 1, # no outline color, scale of distributions
                      alpha = 0.8) + # opacity
  
  # Add point estimates with confidence intervals
  geom_pointinterval(data = forest.data.summary,
                     aes(xmin = .lower, xmax = .upper),
                     fatten_point = 1.5,
                     linewidth = 0.8) +
  
  # Add text and labels
  geom_text(data = mutate_if(forest.data.summary,
                             is.numeric, round, 2), # decimals
            aes(label = glue("{b_Intercept} [{.lower}, {.upper}]"), # format text -> ß [CIlow, CIhigh]
                x = Inf), hjust = "inward") + #aligns text inward from the edge
  labs(x = "beta",
       y = element_blank()) +
  #xlim(-4, 1) +
  theme_light() +
  theme(panel.border = element_blank())
```

If you're wondering why the effect sizes are not the same as those reported in the studies, it's because of a process called _shrinkage_. The Bayesian model estimates 'true' effect sizes for each study and if it detects outliers, they will be pulled towards the overall effect. This is an advantage of the Bayesian approach, particularly when pooling a small number of studies. The model is less sensitive to outliers than frequentist models.
