---
title: "A Bayesian Meta Analysis on Lead Exposure and IQ Loss"
author: "Paulina Sell"
date: "2025-07-25"
output:
    cleanrmd::html_document_clean:
      theme: marx
      toc: TRUE
  # html_document: 
    # theme: paper
    # toc: true
    # toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # global setting to prevent code, but not results from appearing in the finished file (if FALSE)
library(ggplot2)
library(cleanrmd)
```

## Introduction
The aim of this project is to estimate the Environmental Burden of Disease (EBD) attributable to lead exposure in Germany. Here, we will only focus on the neurodevelopmental effects of lead, therefore the health outcome will be IQ point loss in children and adolescents in Germany. In order to estimate these population health effects, we need to find solid evidence for exactly how lead exposure is related to IQ loss. The exposure data we are working with is based on blood tests of ~720 children and adolescents (ages 3 - 17) in Germany from the German Environmental Survey, [GerES V](https://www.umweltbundesamt.de/en/topics/health/assessing-environmentally-related-health-risks/german-environmental-surveys/german-environmental-survey-2014-2017-geres-v#undefined). The data collection period was from 2014 to 2017.\
In order to quantify exactly how lead exposure is related to IQ loss, we consult the epidemiological literature. There are many studies published focusing on the the neurodevelopmental effects of lead. Instead of just choosing one estimate for our EBD analysis, we could try to pool estimates from multiple studies in order to strengthen the evidence we base our analysis on. And since uncertainty plays an important role in the final burden of disease results, we will employ a Bayesian approach. With this approach we naturally model uncertainty and get a full posterior distribution of the pooled effect (instead of a mean effect and an hard-to-interpret confidence interval). Later, we will be able to use many draws from the posterior distribution for our EBD analysis! But first, let's meta analyse some epi studies.


## Step 1: Standardizing Effect Estimates

We first need to search for and identify relevant studies and extract relevant data. These steps were already performed by Philippe Palmont and his team at ANSES.   
Now, we still need to harmonize / standardize the effect estimates before we can pool them. Most of the studies identified transformed their blood lead level (BLL) data to a logarithmic scale in order to fulfill the assumption of a linear relationship between BLL (exposure) and IQ points (outcome). Unfortunately, they did not always use the same bases. Some used log~10~, some ln, and some log~2~. Also, 3 of the 14 included studies did not transform their BLL at all, so it's on a linear scale. Philippe Palmont and his team standardized all estimates to linear increments of exposure in their analysis, using the template proposed by [Rodriguez Barranco 2017](https://doi.org/10.1186/s12874-017-0322-8). But what happens to the effect estimates, when we transform them in this way?

### Visually Checking Transformations

Let's compare the original beta coefficient from Crump et al. 2013 (ß = -2.65) for ln(BLL) with the transformed one (ß = ß -0.277), that now should be used for linear BLL.

```{r echo = F}
beta_log <- -2.650
beta_linear <- -0.277

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_log <- beta_log * log(blood_lead + 1)
IQ_loss_linear <- beta_linear * blood_lead

plot_data <- data.frame(
  blood_lead = blood_lead,
  log_ERF = IQ_loss_log,
  linear_ERF = IQ_loss_linear
)

ggplot(plot_data, aes(x = blood_lead)) +
  geom_line(aes(y = log_ERF, color = "Logarithmic")) +
  geom_line(aes(y = linear_ERF, color = "Linear")) +
  scale_color_manual(values = c("Logarithmic" = "blue", "Linear" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

This looks like we would likely underestimate the effect of lower BLL and overestimate the effect of higher BLL. Since today, BLL are found to be much lower than even 30 µg/dL (in Germany, the range is somewhere from 0.04 to 4.5 µg/dL in children and adolescents), we would likely underestimate the population health effects of BLL on IQ in children. Transforming log estimates to linear seems to be problematic here. 

### Transforming Estimates to ln(BLL)
In order to preserve the correct exposure-response curves and harmonize the effect estimates for our meta analysis, we could try to transform all estimates to ln(BLL), since this closer to the original estimates reported by the studies, since they mostly just used different bases for their logarithmic transformations.   
   
These are the formulas I used:\
$\beta_{ln} = \beta_{linear} \times C_{ref}$\
$\beta_{ln} = \beta_{log_{2}} \div \ln(2)$\
$\beta_{ln} = \beta_{log_{10}} \div \ln(10)$\
\
Where:\
$\beta_{ln}$ = harmonized beta coefficient (natural logarithm)\
$\beta_{linear}$ = original linear beta coefficient\
$\beta_{log_2}$ = original log₂-transformed beta coefficient\
$\beta_{log_{10}}$ = original log₁₀-transformed beta coefficient\
$C_{ref}$ = reference concentration for linear transformation\
\

#### Case 1: log~10~(BLL) -> ln(BLL): Dantzer et al. 2020
```{r echo = F}
Dantzerbeta_log10 <- -13.150
Dantzerbeta_ln <- -5.711

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_log10 <- Dantzerbeta_log10 * log10(blood_lead + 1)
IQ_loss_ln <- Dantzerbeta_ln * log(blood_lead + 1)

Dantzerplot_data <- data.frame(
  blood_lead = blood_lead,
  log_ERF = IQ_loss_log10,
  ln_ERF = IQ_loss_ln
)

ggplot(Dantzerplot_data, aes(x = blood_lead)) +
  geom_line(aes(y = log_ERF, color = "log(10)")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("log(10)" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

We can't even really see the blue curve, so the transformation seems valid.   


#### Case 2: log~2~(BLL) -> ln(BLL): Desrochers-Couture et al. 2018
```{r echo = F}
Desrochersbeta_log2 <- 0.014
Desrochersbeta_ln <- 0.020

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_log2 <- Desrochersbeta_log2 * log2(blood_lead + 1)
IQ_loss_ln <- Desrochersbeta_ln * log(blood_lead + 1)

Desrochersplot_data <- data.frame(
  blood_lead = blood_lead,
  log_ERF = IQ_loss_log2,
  ln_ERF = IQ_loss_ln
)

ggplot(Desrochersplot_data, aes(x = blood_lead)) +
  geom_line(aes(y = log_ERF, color = "log(2)")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("log(2)" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

In this concentration range, we can see that the transformed estimate is a good approximation for the original one.\
**!** Note that this study reported a small protective effect of lead exposure, this is why the curve is flipped.   

#### Case 3: linear(BLL), not transformed -> ln(BLL): Iglesias et al. 2011
In order to transform this estimate, we need to set a reference BLL ($C_{ref}$). Here, I chose 2.2 µg/dL.  
```{r echo = F}
Iglesiasbeta_linear <- -0.940
Iglesiasbeta_ln <- -2.068

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_linear <- Iglesiasbeta_linear * blood_lead
IQ_loss_ln <- Iglesiasbeta_ln * log(blood_lead + 1)

Iglesiasplot_data <- data.frame(
  blood_lead = blood_lead,
  linear_ERF = IQ_loss_linear,
  ln_ERF = IQ_loss_ln
)

ggplot(Iglesiasplot_data, aes(x = blood_lead)) +
  geom_line(aes(y = linear_ERF, color = "linear")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("linear" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

This looks problematic. The concentration range is quit high though, so we might want to zoom in:

```{r echo = F}
blood_lead_low <- seq(0, 5, by = 0.1)

IQ_loss_linear <- Iglesiasbeta_linear * blood_lead_low
IQ_loss_ln <- Iglesiasbeta_ln * log(blood_lead_low + 1)

Iglesiasplot_data <- data.frame(
  blood_lead_low = blood_lead_low,
  linear_ERF = IQ_loss_linear,
  ln_ERF = IQ_loss_ln
)

ggplot(Iglesiasplot_data, aes(x = blood_lead_low)) +
  geom_line(aes(y = linear_ERF, color = "linear")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("linear" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

It looks like, even in the lower concentration range, the transformed estimate is not a good approximation of the original estimate. For my meta analysis, I therefore decided to exclude 3/14 studies with estimates for untransformed BLL data.

## Step 2: Meta Analysis: Setting Priors
Now that we have collected our study data and harmonized the effect estimates, we can finally begin our meta analysis! 

```{r echo = F, message = F}
library(brms)
library(tidyverse)
library(tidybayes)
library(ggridges)
library(glue)
library(extraDistr)
library(bayesplot)
library(HDInterval)

#data = read.csv("data/study_data_leadIQloss.csv")

# try excluding studies that were transformed from linear to log, since transformation may not be valid here
#data <- data[!data$author_year %in% c("Halabicky 2022", "Iglesias 2011", "Min 2009"), ]

# lets look at the data 
#ggplot(data, aes(x = 1:nrow(data), y = beta_ln)) +
  #geom_point() +
  #labs(x = "Study", y = "Main Effect Beta") +
  #theme_minimal()

#ggplot(data, aes(x = 1:nrow(data), y = se_beta_ln)) +
 # geom_point() +
  #labs(x = "Study", y = "Heterogeneity Tau") +
  #theme_minimal()
```

