---
title: "PARC T6.2.4 – Case study: Frequentist & Bayesian Meta Analysis on Lead Exposure & IQ Loss"
author: "Paulina Sell, UBA"
date-modified: "2025-11-03"
format:
  html:
    toc: true
    self-contained: true
    embed-resources: true
    theme: cerulean
    page-layout: full
fig-align: center
fig-height: 4
---

```{r}
#| label: loading-data
#| echo: false

data_clean <- read.csv("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/data/study_data_leadIQloss_clean.csv")
rationale <- read.csv("/Users/paulinasell/Documents/UBA/PARC/Metaanalysis_lead_IQloss/RProj/data/Rationale_ex-inclusion.csv")
```

# Introduction
This document describes the methodology of a frequentist and a Bayesian meta analysis on the effects of blood lead levels (BLL) in children (postnatal & concurrent) and their intellectual development (IQ). 
The results of these meta analyses will be used in burden of disease estimation within this methodological case study. 
The literature review was conducted by ANSES. 
\

# Step 1: Harmonizing Effect Estimates

In order to harmonize the effect estimates before we can pool them, we need to transform some of them. 
Most of the studies transformed their blood lead level (BLL) data to a logarithmic scale in order to fulfill the assumption of a linear relationship between BLL (exposure) and IQ points (outcome) in their regression analyses. 
They did not always use the same log base. 
Some used log~10~, some ln, and some used log~2~. 
Also, some of the included studies did not transform their BLL at all, so it is on a linear scale. 
ANSES standardized all estimates to linear increments of BLL, using the template published by [Rodriguez Barranco 2017](https://doi.org/10.1186/s12874-017-0322-8). 
The following visual tests will explain the rationale behind our decision to transform all effect estimates to a logarithmic scale.\

## Visualizing Transformations

Comparison of original beta coefficient from [Crump et al. 2013](https://doi.org/10.3109/10408444.2013.832726) (ß = -2.65) for ln(BLL) to the transformed one (ß = -0.277) (now for linear BLL).

```{r}
#| label: lin-vs-log
#| echo: false

library(ggplot2)

beta_log <- -2.650
beta_linear <- -0.277

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_log <- beta_log * log(blood_lead + 1)
IQ_loss_linear <- beta_linear * blood_lead

plot_data <- data.frame(
  blood_lead = blood_lead,
  log_ERF = IQ_loss_log,
  linear_ERF = IQ_loss_linear
)

ggplot(plot_data, aes(x = blood_lead)) +
  geom_line(aes(y = log_ERF, color = "Logarithmic")) +
  geom_line(aes(y = linear_ERF, color = "Linear")) +
  scale_color_manual(values = c("Logarithmic" = "blue", "Linear" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

As visible in the graph, the effects of lower BLL would likely be underestimated in lower BLL ranges and overestimated in higher BLL ranges.
Since today, BLL are often found to be very low (range in GerES V: approximately 0.04 to 4.5 µg/dL), the population health effects (Burden of Disease) of BLL would likely be underestimated.
Therefore, effect estimates were transformed to ln(BLL) when necessary in order to harmonize them. 

Formulas used to transform effect estimates to ln(BLL):\
$$
\beta_{ln} = \beta_{linear} \times C_{ref}\
$$
$$
\beta_{ln} = \frac{beta_{log_{2}}} {\ln(2)}\
$$

$$
\beta_{ln} = \frac{beta_{log_{10}}} {\ln(10)}\
$$
Where:\
$\beta_{ln}$ = harmonized beta coefficient (natural logarithm)\
$\beta_{linear}$ = beta coefficient for linear BLL\
$\beta_{log_2}$ = beta coefficient for log₂-transformed BLL\
$\beta_{log_{10}}$ = beta coefficient for log₁₀-transformed BLL\
$C_{ref}$ = reference concentration for linear transformation\

In the following section, all three cases of transformations are visualized in order to check for validity.

### Case 1: log~10~(BLL) to ln(BLL): [Dantzer et al. 2020](https://doi.org/10.1016/j.scitotenv.2019.134519)
```{r}
#| label: log10-to-ln
#| echo: false


Dantzerbeta_log10 <- -13.150
Dantzerbeta_ln <- -5.711

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_log10 <- Dantzerbeta_log10 * log10(blood_lead + 1)
IQ_loss_ln <- Dantzerbeta_ln * log(blood_lead + 1)

Dantzerplot_data <- data.frame(
  blood_lead = blood_lead,
  log_ERF = IQ_loss_log10,
  ln_ERF = IQ_loss_ln
)

ggplot(Dantzerplot_data, aes(x = blood_lead)) +
  geom_line(aes(y = log_ERF, color = "log(10)")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("log(10)" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

The transformation from log~10~(BLL) to ln(BLL) does not affect the exposure-response curve.


### Case 2: log~2~(BLL) to ln(BLL): [Desrochers-Couture et al. 2018](https://doi.org/10.1016/j.envint.2018.10.043)
```{r}
#| label: log2-to-ln
#| echo: false



Desrochersbeta_log2 <- 0.014
Desrochersbeta_ln <- 0.020

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_log2 <- Desrochersbeta_log2 * log2(blood_lead + 1)
IQ_loss_ln <- Desrochersbeta_ln * log(blood_lead + 1)

Desrochersplot_data <- data.frame(
  blood_lead = blood_lead,
  log_ERF = IQ_loss_log2,
  ln_ERF = IQ_loss_ln
)

ggplot(Desrochersplot_data, aes(x = blood_lead)) +
  geom_line(aes(y = log_ERF, color = "log(2)")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("log(2)" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

The transformation from log~2~(BLL) to ln(BLL) does not affect the exposure-response curve in the lower exposure ranges.
Note: this study reported a small protective effect of lead exposure, this is why the curve is flipped.   

### Case 3: linear(BLL) to ln(BLL): [Iglesias et al. 2011](https://doi.org/10.1179/107735211799041841)
In order to transform this estimate, we need to set a reference BLL ($C_{ref}$). 2.2 µg/dL was set in this transformation.  
```{r}
#| label: lin-to-ln
#| echo: false


Iglesiasbeta_linear <- -0.940
Iglesiasbeta_ln <- -2.068

blood_lead <- seq(0, 50, by = 0.1)

IQ_loss_linear <- Iglesiasbeta_linear * blood_lead
IQ_loss_ln <- Iglesiasbeta_ln * log(blood_lead + 1)

Iglesiasplot_data <- data.frame(
  blood_lead = blood_lead,
  linear_ERF = IQ_loss_linear,
  ln_ERF = IQ_loss_ln
)

ggplot(Iglesiasplot_data, aes(x = blood_lead)) +
  geom_line(aes(y = linear_ERF, color = "linear")) +
  geom_line(aes(y = ln_ERF, color = "ln")) +
  scale_color_manual(values = c("linear" = "blue", "ln" = "red")) +
  labs(
    x = "Blood Lead Level (µg/dL)",
    y = "IQ Loss (points)"
  ) +
  theme_minimal()
```

It appears that the transformed estimate does not accurately approximate the original estimate. 
For this meta analysis, we therefore decided to exclude all studies with estimates for untransformed BLL data.\

\
These are the final studies and transformed effect estimates that we will be working with.
```{r}
#| label: view-data
#| echo: false

knitr::kable(
  rationale[,!(names(rationale) %in% "doi")],
  col.names = c('Author & Year', 'N', 'Beta', 'SE(Beta)', 'Original BLL Sclae', 'Incl.  Analysis 1', 'Incl. Analysis 2', 'Comment'),
  digits = 2)
```
